General purpose gene annotation using deep learning

### 2020-01-07 paper/main.tex
paper

\section{Approach}
Helixer is a prototype software for training and utilizing a general purpose gene annotation model for eukaryotic genomes. We demonstrate the effectiveness of this approach by training a single model for animals (phylogenetic names?) and plants respectively, which was then evaluated on 192 genomes in the animal case and 60 genomes for the plant model. We leverage modelling techniques from the field of deep learning, which has been shown to have the potential to be superior in flexibility and accuracy to traditional machine learning methods in a variety of fields. Our model expects the raw genomic sequence as main input and produces functional predictions on a base pair wise level.

@@

The selection of training genomes required balancing multiple conflicting tradeoffs. On the one hand, we do want to train with as much data as possible but not all data has a good enough quality to enable a powerful generalization. It is also very desirable to have a broad spread in phylogenetic distance as well as genome size and average gene length as the model is increasingly unlikely to generalize well beyond the borders given by the training data. It was also important to us to get experimental results within a couple of days as we tested many different data and model configurations.

\begin{table}[!t]
\processtable{Data arrays generated and used by Helixer\label{Tab:01}} {
\begin{tabular}{@{}llll@{}}
\toprule Name & Information \\
\midrule
Input & Genomic sequence in a one hot encoding\\
Output & Labels in a 4-dimensional one hot encoding \\
Sample weights & One of $\{0,1\}$; whether there is an error at a base\\
Gene lengths & The length of a gene at a base or 0 if intergenic \\
RNASeq score & A score in the interval $...$ to rank the coherence of the \\
		     & annotation with RNASeq information \\
\botrule
\end{tabular}}{}
\end{table}

>>>>

\section{Approach}
Helixer is a prototype software for training and utilizing a general purpose gene annotation model for eukaryotic genomes. We demonstrate the effectiveness of this approach by training a single model for animals (phylogenetic names?) and plants respectively, which was then evaluated on 192 genomes in the animal case and 60 genomes for the plant model. We leverage modelling techniques from the field of deep learning, which has been shown to have the potential to be superior in flexibility and accuracy to traditional machine learning methods in a variety of fields. Our model expects the raw genomic sequence as main input and produces functional predictions on a base pair wise level. The source code and all data is publically available.

@@

The selection of training genomes required balancing multiple conflicting tradeoffs. On the one hand, we do want to train with as much data as possible, but not all data has a good enough quality to enable a powerful generalization. It is also very desirable to have a broad spread in phylogenetic distance as well as genome size and average gene length as the model is increasingly unlikely to generalize well beyond the borders given by the training data. It was also important to us to get experimental results within a couple of days as we tested many different data and model configurations, which limited the overall size of the training set and the model.

\subsection{Model Architecture}
	We use a 4-layer deep stacked LSTM network with 256 units per layer to produce the predictions. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras (citation) on top of TensorFlow (citation). All relevant hyperparameters can be found in Table \ref{Tab:02} or in the Helixer source code repository.

The model was trained with 10 bases of genomic sequence as input during each time step and produces individual predictions for each of those 10 bases simultaneously. This grouping enabled us to train with far longer sequences than normally done.

\subsection{Evaluation}

\begin{table}[!t]
\processtable{Data arrays generated and used by Helixer\label{Tab:01}} {
\begin{tabular}{@{}ll@{}}
\toprule Name & Information \\
\midrule
Input & Genomic sequence in a one hot encoding\\
Output & Labels in a 4-dimensional one hot encoding \\
Sample weights & One of $\{0,1\}$; whether there is an error at a base\\
Gene lengths & The length of a gene at a base or 0 if intergenic \\
RNASeq score & A score in the interval $...$ to rank the coherence of the \\
		     & annotation with RNASeq information \\
\botrule
\end{tabular}}{}
\end{table}

\begin{table}[!t]
\processtable{Hyperparameters\label{Tab:02}} {
\begin{tabular}{@{}ll@{}}
\toprule Name & Value \\
\midrule
Layers & 4\\
Units per layer & 256\\
Learning rate & 1e-4\\
Optimizer & Adam\\
\botrule
\end{tabular}}{}
\end{table}
